Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-12 paper/main.tex
draft

@@

\abstract{\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/HelixerPrep} [change] under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{alisandra.denton@hhu.de}{alisandra.denton@hhu.de}\\  
% normally contact is last author / person with permanent position (i.e. Andreas). 
% Not sure how important that is, but if we're skipping that, then we can change it back to Felix
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.}

@@

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant
species from the JGI Phytosome 13 database \citep{goodstein2012phytozome}; exact 
genomes used are list here \ref{}. % add a supplemental table with names and versions
Data was downloaded on 29/03/2019 and 15/10/2019 for plants and animals respectively. 
Both data groups were used seperately during training as well as during evaluation.

@@

\subsection{Evaluation of AUGUSTUS} 
\label{sec:augustus}
What Marvin did, why we don't use Genic F1 for comparison, etc

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{tab:arrays}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in the 4-dimensional IUPAC encoding\\
	  & (one hot encoding for non-ambiguous bases) \\
Output & Labels in a 4-dimensional one hot encoding representing \\
      & the classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}\\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{{\it Note:} An intron inside a UTR region is encoded as intron and 
can be inferred from the prediction order.}
\end{table}

@@

Eliminating the need for a close genome annotation being present for pretraining.

@@

\section*{Acknowledgements}

We are very grateful for the support of...

\section*{Funding}

This work has been supported by the... \vspace*{-12pt}

>>>>

\usepackage{draftwatermark}
\SetWatermarkText{DRAFT}
\SetWatermarkScale{1}

@@

\abstract{\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/HelixerPrep} under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{alisandra.denton@hhu.de}{alisandra.denton@hhu.de}\\  
% normally contact is last author / person with permanent position (i.e. Andreas). 
% Not sure how important that is, but if we're skipping that, then we can change it back to Felix
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.}

@@

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{tab:arrays}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in the 4-dimensional IUPAC encoding\\
	  & (one hot encoding for non-ambiguous bases) \\
Output & Labels in a 4-dimensional one hot encoding representing \\
      & the classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}\\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{{\it Note:} An intron inside a UTR region is encoded as intron and 
can be inferred from the prediction order.}
\end{table}
\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant
species from the JGI Phytosome 13 database \citep{goodstein2012phytozome}; exact 
genomes used are list here \ref{}. % add a supplemental table with names and versions
Data was downloaded on 29/03/2019 and 15/10/2019 for plants and animals respectively. 
Both data groups were used seperately during training as well as during evaluation.

@@

@@

@@
